Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes

### 2019-05-07 GWLDCI.tex
put in one figure
put in the rest of the associahedron figures

It turns out that this defines a class of combinatorially equivalent polytopes, which can be proved by giving a realization.  There are many interesting realizations of the associahedron \cite{CSZinequivalent}; the following will be the one which we will use. \sanote{picture of an associahedron somewhere?}\note{Yes, I agree that there should be a picture}

@@

The map $\tau$ defined in Definition \ref{WLDtriangulationdfn} assigns a dissection of an $n$-gon to a Wilson loop diagram $W = (\cP, [n])$. Recall that if $|\cP| = k$, then $\tau(W)$ is a dissection with $k$ chords. Therefore, the map $\tau$ associates to each weakly admissible Wilson loop diagram an $n-3 - k$ dimensional face of the associahedron $A_n$. \sanote{picture of $\tau(W)$ for $k =1$, $n = 5$?}

@@

  Consider now an edge whose quadrilateral has corners $x_i$, $x_j$, $x_k$, and $x_l$ in that order.  Let $t_1$ and $t_2$ be the triangulations corresponding to the two ends of the edge where $t_1$ has a diagonal from $x_j$ to $x_l$ and $t_2$ has a diagonal from $x_i$ to $x_k$.  From the definition of $A_n$ we see that $s_{t_1}$ and $s_{t_2}$ only differ at coordinates $i$, $j$, $k$, and $l$, and all of the triangles incident to any of these vertices other than the one coming from triangulating $x_i, x_j, x_k, x_l$ are the same in both $t_1$ and $t_2$.  Let $a$ be the area of the quadrilateral and let $a(i,j,k)$ be the area of the triangle with corners $x_i$, $x_j$ and $x_k$, and similarly for other indices then $s_{t_1}$ on coordinates $i$, $j$, $k$, $l$ is $(a(i,j,l) + b_i, a+b_j, a(k,l,j)+b_k, a+b_l)$ and $s_{t_2}$ on these coordinates is $(a + b_i, a(j,k,i)+b_j, a+b_k, a(l,i,k)+b_l$, where the $b_i$ are the sums of the areas of the relevant triangles which are common to both triangulations.  Thus a vector parallel to this edge is the vector which is $(a(i,j,l) - a, a-a(j,k,i), a(k,l,j)-a, a-a(l,i,k)) = (-a(j,k,l), a(i,k,l), -a(i,j,l), a(i,j,k))$ on $i$, $j$, $k$, $l$ and $0$ in all other coordinates. \sanote{definitely need a picture}

@@

\begin{lem}\label{lem good quads}
  Let $D_1$ and $D_2$ be two dissections of the convex polygon defined by the vertices $x_1, \ldots, x_n$.  Suppose $V_{>3}(D_1) = V_{>3}(D_2)$, but that the $D_1$ and $D_2$ do not have the same set of non-triangle faces.  Then there are four vertices $x_i$, $x_j$, $x_k$, $x_l$ such that in one of $D_1$ or $D_2$ all four vertices are on the same non-triangle face while in the other there is a diagonal of the dissection such that there is one side of the diagonal with exactly one of $\{x_i, x_j, x_k, x_l\}$ (so the others are end points of this diagonal or strictly on the other side).
\end{lem} \sanote{picture?}

@@

Assume for a contradiction that no quadrilateral as in the statement exists, and let $d$ and $f$ be as above.  Then $f$ must have two vertices strictly on each side of $d$, say $x_i$ and $x_j$ on one side and $x_k$, $x_l$ on the other side (and in that order).  Consider the face incident to $d$ on the $x_i, x_j$ side of $d$.  Suppose this face is a triangle, then its other vertex cannot be in $[x_i, x_j]$ as otherwise taking one of the other edges of the triangles would give a configuration as in the statement.  Suppose now that this face is a quadrilateral, again no other vertex of this face can be in $(x_i, x_j)$ as otherwise again taking one of the other edges of the face as $d$ we'd get a quadrilateral as in the statement, the same holds if one of the other vertices of the face is $x_i$ or $x_j$ but the other is not.  If the two other vertices of the face are $x_i$ and $x_j$, then we also get a configuration as in the statement but now with the four vertices being the two ends of $q$ and $x_i, x_j$ and with one of the edges of $f$ as the diagonal.  Thus we can conclude that none of the vertices of this face are in $[x_i, x_j]$.  Also if the face has three vertices on one side of $f$ and one on the other then we get a quadrilateral as in the statement again with this face giving the four vertices and the diagonal is a side of $f$.  In this case the face is either a triangle with its third vertex of the triangle must be in $(x_j,x_k)$ or in $(x_l,x_i)$ or a quadrilateral with one more vertex in $(x_j, x_k)$ and one more vertex in $(x_l, x_i)$.  Taking all this together, whether the face is a triangle or a quadrilateral, replacing $d$ with one of the other edges of the face we get a new diagonal $d'$ which also separates $f$ into $x_i, x_j$ and $x_k, x_l$ but is strictly closer to $[x_i, x_j]$ than $d$ was.  The same argument holds for $d'$, but the polygon is finite, so this process cannot go on indefinitely, contradicting that no quadtilateral as in the statement exists.
\note{That huge paragraph also needs a picture}
\end{proof}

>>>>

It turns out that this defines a class of combinatorially equivalent polytopes, which can be proved by giving a realization.  There are many interesting realizations of the associahedron \cite{CSZinequivalent}; the following will be the one which we will use.  See Figure~\ref{associahedron5gon} for examples.

@@

The map $\tau$ defined in Definition \ref{WLDtriangulationdfn} assigns a dissection of an $n$-gon to a Wilson loop diagram $W = (\cP, [n])$. Recall that if $|\cP| = k$, then $\tau(W)$ is a dissection with $k$ chords. Therefore, the map $\tau$ associates to each weakly admissible Wilson loop diagram an $n-3 - k$ dimensional face of the associahedron $A_n$. This is illustrated for $n=5$ on the left in Figure~\ref{associahedron5gon}.

\begin{figure}
\includegraphics{associahedron5gon} \hspace{1cm} \includegraphics{associahedron3d}
\caption{Two associahedra labelled by all and some (respectively) of their polygon dissections.}\label{associahedron5gon}
\end{figure}

@@

  Consider now an edge whose quadrilateral has corners $x_i$, $x_j$, $x_k$, and $x_l$ in that order.  Let $t_1$ and $t_2$ be the triangulations corresponding to the two ends of the edge where $t_1$ has a diagonal from $x_j$ to $x_l$ and $t_2$ has a diagonal from $x_i$ to $x_k$.  From the definition of $A_n$ we see that $s_{t_1}$ and $s_{t_2}$ only differ at coordinates $i$, $j$, $k$, and $l$, and all of the triangles incident to any of these vertices other than the one coming from triangulating $x_i, x_j, x_k, x_l$ are the same in both $t_1$ and $t_2$.  Let $a$ be the area of the quadrilateral and let $a(i,j,k)$ be the area of the triangle with corners $x_i$, $x_j$ and $x_k$, and similarly for other indices then $s_{t_1}$ on coordinates $i$, $j$, $k$, $l$ is $(a(i,j,l) + b_i, a+b_j, a(k,l,j)+b_k, a+b_l)$ and $s_{t_2}$ on these coordinates is $(a + b_i, a(j,k,i)+b_j, a+b_k, a(l,i,k)+b_l$, where the $b_i$ are the sums of the areas of the relevant triangles which are common to both triangulations.  Thus a vector parallel to this edge is the vector which is $(a(i,j,l) - a, a-a(j,k,i), a(k,l,j)-a, a-a(l,i,k)) = (-a(j,k,l), a(i,k,l), -a(i,j,l), a(i,j,k))$ on $i$, $j$, $k$, $l$ and $0$ in all other coordinates.  This is illustrated in Figure~\ref{triangulate}.

\begin{figure}
\includegraphics{triangulate} 
\caption{The shaded regions are triangulated regions which are the same in $t_1$ and $t_2$.  Consider $s_{t_1}$ and $s_{t_2}$ at coordinate $i$.  In $s_{t_1}$ this coordinate is given by the area of those shaded triangles incident to $x_i$ along with $a(i,j,l)$.  In $s_{t_2}$ this coordinate is given by the area of those shaded triangles incident to $x_i$ along with $a(i,l,k) + a(i,j,k) = a$.}\label{triangulate}
\end{figure}

@@

\begin{lem}\label{lem good quads}
  Let $D_1$ and $D_2$ be two dissections of the convex polygon defined by the vertices $x_1, \ldots, x_n$.  Suppose $V_{>3}(D_1) = V_{>3}(D_2)$, but that the $D_1$ and $D_2$ do not have the same set of non-triangle faces.  Then there are four vertices $x_i$, $x_j$, $x_k$, $x_l$ such that in one of $D_1$ or $D_2$ all four vertices are on the same non-triangle face while in the other there is a diagonal of the dissection such that there is one side of the diagonal with exactly one of $\{x_i, x_j, x_k, x_l\}$ (so the others are end points of this diagonal or strictly on the other side).
\end{lem}

The situation described in the lemma is illustrated in Figure~\ref{configs}.

\begin{figure}
\includegraphics{configs} 
\caption{Up to reflection and to rotation of $x_i$, $x_j$, $x_k$, $x_l$, the possible configurations of the quadrilateral from one dissection and diagonal from the other dissection (thick edge) illustrated together on the same vertex set.}\label{configs}
\end{figure}

@@

Assume for a contradiction that no quadrilateral as in the statement exists, and let $d$ and $f$ be as above.  Then $f$ must have two vertices strictly on each side of $d$, say $x_i$ and $x_j$ on one side and $x_k$, $x_l$ on the other side (and in that order).  Consider the face incident to $d$ on the $x_i, x_j$ side of $d$.  Suppose this face is a triangle, then its other vertex cannot be in $[x_i, x_j]$ as otherwise taking one of the other edges of the triangles would give a configuration as in the statement.  Suppose now that this face is a quadrilateral, again no other vertex of this face can be in $(x_i, x_j)$ as otherwise again taking one of the other edges of the face as $d$ we'd get a quadrilateral as in the statement, the same holds if one of the other vertices of the face is $x_i$ or $x_j$ but the other is not.  If the two other vertices of the face are $x_i$ and $x_j$, then we also get a configuration as in the statement but now with the four vertices being the two ends of $q$ and $x_i, x_j$ and with one of the edges of $f$ as the diagonal.  Thus we can conclude that none of the vertices of this face are in $[x_i, x_j]$.  Also if the face has three vertices on one side of $f$ and one on the other then we get a quadrilateral as in the statement again with this face giving the four vertices and the diagonal is a side of $f$.  In this case the face is either a triangle with its third vertex of the triangle must be in $(x_j,x_k)$ or in $(x_l,x_i)$ or a quadrilateral with one more vertex in $(x_j, x_k)$ and one more vertex in $(x_l, x_i)$.  The three possibilities are illustrated in Figure~\ref{inductive_step}. Taking all this together, whether the face is a triangle or a quadrilateral, replacing $d$ with one of the other edges of the face we get a new diagonal $d'$ which also separates $f$ into $x_i, x_j$ and $x_k, x_l$ but is strictly closer to $[x_i, x_j]$ than $d$ was.  The same argument holds for $d'$, but the polygon is finite, so this process cannot go on indefinitely, contradicting that no quadtilateral as in the statement exists.
\end{proof}

\begin{figure}
\includegraphics{inductive_step} 
\caption{The three possible configurations in the final part of the proof of Lemma~\ref{lem good quads}.  The thin lines come from the polygon dissection with $f$ as a face while the thick lines come from the polygon dissection with $d$ as an edge.}\label{inductive_step}
\end{figure}
